Estimating Predictive Rate-Distortion Curves via Neural Variational Inference

### 2019-04-18 paper.tex

\newif \ifcomment
\commentfalse

@@

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\infdiv}{D\infdivx}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}

@@

@@

In Section~\ref{sec:background}, we formally introduce predictive rate distortion, and discuss related notions of predictive complexity.
In Section~\ref{sec:oce}, we describe the prior method of Optimal Causel Estimation (OCE).
In Section~\ref{sec:nprd}, we describe our method NPRD.
In Section~\ref{sec:experiments}, we validate NPRD on processes whose predictive rate distortion is analytically known, showing that it finds essentially optimal predictive models.
In Section~\ref{sec:language}, we apply NPRD to data from five languages, providing the first estimates of predictive rate distortion of natural language.

@@

We consider stationary stochastic processes $(X_t)_{t \in \mathbb{Z}}$, taking values in a state space $S$.
Given a reference point in time, say $T=0$, we are interested in the problem of predicting the future of $X_{t\geq 0} = (X_0, X_1, X_2, ...)$ from the past $X_{t< 0} = (..., X_{-2}, X_{-1})$.
In general -- unless the observations $X_t$ are independent -- predicting the future of the process accuractely will require taking into account the past observations.
There is a tradeoff between the accuracy of prediction, and how much information about the past is being taken into account.
On one extreme, not taking the past into account at all, one will not be able to take advantage of the dependencies between past and future observations.
On the other extreme, considering the entirety of the past observations $X_{t \leq 0}$ can require storing large amounts of information.
This tradeoff is referred to as \emph{Predictive Rate-Distortion} \citep{marzen-predictive-2016}.
The term \emph{rate} refers to the amount of past information being taken into account, while \emph{distortion} refers to the degradation in prediction compared to optimal prediction from the full past.

The problem of predictive rate-distortion has been formalized by a range of studies.
A principled and general formalization is provided by applying the Information Bottleneck idea \citep{tishby-information-1999, still-optimal-2010,marzen-predictive-2016}:
We will write $\overleftarrow{X}$ for the past $X_{< 0}$, and $\overrightarrow{X}$ for the future $X_{\geq 0}$, following \cite{marzen-predictive-2016}.
We consider random variables $Z$, called \emph{codes}, that summarize the past and are used by the observer to predict the future.
Formally, $Z$ needs to be independent of the future $\overrightarrow{X}$ conditional on the past $\overleftarrow{X}$. Symbolically, 
\begin{equation}\label{eq:markov}
	Z \bot \overrightarrow{X} | \overleftarrow{X}
\end{equation}
This is equivalent to the requirement that $Z \rightarrow \overleftarrow{X} \rightarrow\overrightarrow{X}$ be a Markov chain.
This formalizes the idea that the code is computed by an observer from the past, without having access to the future.
Predictions are then made based only on $Z$, without additional access to the past $\overleftarrow{X}$.

The \emph{rate} of the code $Z$ is the mutual information between $Z$ and the past: $\operatorname{I}[Z, \overleftarrow{X}]$.
By the Channel Coding Theorem, this describes the channel capacity that the observer requires in order to transform past observations into the code $Z$.

The \emph{distortion} is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\overleftarrow{X}$.
In the Information Bottleneck formalization, this
is equivalent to the amount of mutual information between past and future that is not captured by $Z$:
\begin{equation}
	\operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z]
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	\operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z] = \operatorname{I}[\overleftarrow{X}, \overrightarrow{X}] - \operatorname{I}[Z, \overrightarrow{X}]
\end{equation}
i.e., it captures how much less information $Z$ carries about the future $\overrightarrow{X}$, compared to the full past $\overleftarrow{X}$.
For a fixed process $(X_t)_t$, choosing $Z$ to minimize the distortion is equivalent to maximizing the mutual information between the code and the future:
\begin{equation}\label{eq:predictiveness}
\operatorname{I}[Z, \overrightarrow{X}]
\end{equation}
We will refer to~(\ref{eq:predictiveness}) as the \emph{predictiveness} of the code $Z$.

The rate-distortion tradeoff then chooses $Z$ to %minimize rate at bounded distortion:
%$$\min_{Z : \operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z] \leq d} \operatorname{I}[Z, \overleftarrow{X}]$$
%, or -- equivalently --
 minimize distortion at bounded rate:
\begin{equation}
\min_{Z :  \operatorname{I}[\overleftarrow{X}, Z] \leq d}  \operatorname{I}[\overleftarrow{X}, \overrightarrow{X}|Z] 
\end{equation}
or -- equivalently -- maximize predictiveness at bounded rate:
\begin{equation}
\max_{Z :  \operatorname{I}[\overleftarrow{X}, Z] \leq d}  \operatorname{I}[Z, \overrightarrow{X}] 
\end{equation}
Equivalently, for each  $\lambda \geq 0$, we study the problem 
\begin{equation}\label{eq:ib}
	\max_{Z} \left( \operatorname{I}[Z, \overrightarrow{X}] - \lambda \cdot \operatorname{I}[ \overleftarrow{X}, Z] \right)
\end{equation}
where the scope of the maximization is the class of all random variables $Z$ such that $Z \rightarrow \overleftarrow{X} \rightarrow \overrightarrow{X}$ is a Markov chain. %, that is, $Z$ is independent of the future given the past.

The objective (\ref{eq:ib}) is equivalent to the Information Bottleneck \citep{tishby-information-1999}, applied to the past and future of a stochastic process.
The coefficient $\lambda$ indicates how strongly a high rate $\operatorname{I}[\overleftarrow{X}, Z]$ is penalized; higher values of $\lambda$ result in lower rates and thus lower values of predictiveness.

The largest achievable predictiveness $\operatorname{I}[Z, \overrightarrow{X}]$ is equal to $\operatorname{I}[\overleftarrow{X}, \overrightarrow{X}]$, which is known as the \emph{excess entropy} of the process~\citep{feldman-synchronizing-2004}.
Due to the Markov condition~(\ref{eq:markov}) and the Data Processing Inequality, predictiveness of a code $Z$ is always upper-bounded by the rate:
\begin{equation}
\operatorname{I}[Z, \overrightarrow{X}] \leq \operatorname{I}[\overleftarrow{X}, Z] 
\end{equation}
As a consequence, when $\lambda \geq 1$, then (\ref{eq:ib}) is always optimized by a trivial $Z$ with zero rate and zero predictiveness.
When $\lambda = 0$, any lossless code optimizes the problem.
Therefore, we will be concerned with the situation where $\lambda \in (0,1)$.

%
%
%Predictive rate-distortion describes the tradeoff between the cost of representing features of the past, and the quality of predictions in the future.
%Let $(X_t)$ be a stationary process.
%We consider feature representations $Z$ of the past $\overleftarrow{X}$ that are used to predict the future $\overrightarrow{X}$.
%There is a tradeoff between the complexity of $Z$ and the success of predicting the future:
%More complex representations of the past can reduce future prediction loss -- up to a certain limit given by the entropy rate of the process. %This limit corresponds to perfect modeling, and results in Crutchfield and colleagues' notion of complexity.
%
%The complexity of the representation $Z$, or the \emph{rate}, is formalized as the mutual information between the past and the representation $Z$: $\operatorname{I}[Z, \overleftarrow{X}]$.
%Different distortion measures could be used as measures of prediction loss.
%CITE maximize the entropy, or, equivalently, minimize the mutual information, between $Z$ and the future: $\operatorname{I}[Z, \overleftarrow{X}]$.
%
%Thus, for any process $(X_t)$, we obtain a rate-distortion tradeoff curve.
%The achievable region is populated by the random variables $Z$ such that $Z \rightarrow \left(\overleftarrow{X}\right) \rightarrow \left(\overrightarrow{X}\right)$ is a Markov chain.
%The points on the tradeoff curve are obtained by solving, for each $\beta \in [0,1]$, the problem
%\begin{equation}\label{eq:ib}
%	\max_{Z} \left( \operatorname{I}[\overrightarrow{X}, Z] - \beta \cdot \operatorname{I}[Z, (\overleftarrow{X})] \right)
%\end{equation}
%where the scope of the maximization is the class of all random variables $Z$ such that $Z - \left(\overleftarrow{X}\right) - \left(\overrightarrow{X}\right)$ is a Markov chain, that is, $Z$ is independent of the future given the past.
%This is exactly the Information Bottleneck \cite{tishby-information-1999}, applied to the past and future of a stochastic process.
%
%

@@

Predictive Rate-Distortion is closely related to Statistical Complexity and the $\epsilon$-machine \citep{crutchfield-inferring-1989}.
Given a stationary process $X_t$, its \emph{causal states} are the equivalence classes of semi-infinite pasts $\overleftarrow{X}$ that induce the same conditional probability over semi-infinite futures $\overrightarrow{X}$:
Two pasts $\overleftarrow{x}, \overleftarrow{x}'$ belong to the same causal state if and only if $P(x_{1...M}|\overleftarrow{x}) = P(x_{1...M}|\overleftarrow{x}')$ holds for all finite sequences $x_{0...M}$ ($M \in \mathbb{N}$).
Note that this definition is not measure-theoretically rigorous; such a treatment is provided by~\citet{lohr-properties-2009}.

The causal states constitute the state set of a a Hidden Markov Model (HMM) for the process, referred to as the \emph{$\epsilon$-machine}  \citep{crutchfield-inferring-1989}.
The \emph{Statistical Complexity} of a process is the state entropy of the $\epsilon$-machine.
Statistical complexity can be computed easily if the $\epsilon$-machine is analytically known, but estimating statistical complexity empirically from time series data is very challenging and seems to at least require additional assumptions about the process~\citep{clarke2003application}.

@@

There are related models that represent past observations by extracting those features that are relevant for prediction.
Predictive State Representations \citep{singh-learning-2003,singh-predictive-2004} and Observable Operator Models \citep{jaeger1998discrete} encode past observations as sets of predictions about future observations.
\citet{rubin2012trading} study agents that trade the cost of representing information about the evironment against the reward they can obtain by choosing actions based on the representation.
Relatedly, \citet{still-information-2014} introduces a Recursive Past Future Information Bottleneck where past information is compressed repeatedly, not just at one reference point in time.

As discussed in Section~\ref{sec:complexity}, estimating predictive rate-distortion is related to the problem of estimating statistical complexity.
\citet{clarke2003application} and \citet{still-optimal-2010} consider the problem of estimating statistical complexity from finite data.
While statistical complexity cannot be identified from finite data in general \citep{clarke2003application}, \citet{clarke2003application} introduce certain restrictions on the underlying process that make this more feasible.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Estimation (OCE, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts $(X_{-M \dots -1}), (X_{0 \dots M-1})$:
\begin{equation}\label{eq:ib-oce}
	\max_{Z} \left( \operatorname{I}[X_{-M\dots -1}, Z] - \lambda \cdot \operatorname{I}[Z, X_{0\dots M-1}] \right)
\end{equation}
Second, it estimates information measures directly from the observed counts of $(X_{-M, -1}), (X_{0, M-1})$ using the plug-in estimator of mutual information.
With such an estimator, the problem in (\ref{eq:ib}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|X_{-M...-1})$ that maps each observed past sequence $X_{-M...-1}$ to a distribution over a (finite) set of code words $Z$.
%From this, one obtains a corresponding decoder $P(X_{0...M}|Z)$ via Bayes' Rule.

Two main challenges have been noted in prior work:
First, solving the problem for a finite empirical sample leads to overfitting, overestimating the amount of structure in the process.
\citet{still-optimal-2010} address this by subtracting an asymptotic correction term that becomes valid in the limit of large $M$ and $\lambda \rightarrow 0$, when the codebook $P(Z|\overleftarrow{X})$ becomes deterministic, and which allows them to select a deterministic codebook of an appropriate complexity.
This leaves open how to obtain estimates outside of this regime, when the codebook can be far from deterministic.

@@

A further challenge is that OCE is infeasible if the number of required codewords is too large, again because OCE requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $\operatorname{I}[\overleftarrow{X}, Z] > log(10^5) \approx 11.5$ cannot be captured with OCE.

@@

We now introduce our method, Neural Predictive Rate Distortion (NPRD), to address the shortcomings of OCE, by using parametric function approximation:
Whereas OCE constructs a codebook mapping between observed sequences and codes, we use general-purpose function approximation estimation methods to compute the representation $Z$ from the past and to estimate a distribution over future sequences from $Z$.
In particular, we will use recurrent neural networks, which are known to provide good models of sequences from a wide range of domains; our method will be applicable to other types of function approximators also.
%The parameters of the function approximators are optimized on the observed pairs of pasts and futures.
%When deploying the resulting mapping on unseen data, the function approximator can be used to generate appropriate representations $Z$ even for sequences that did not occur in the initial data sample.

@@

As in OCE (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures:
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}} \left( \operatorname{I}[Z, X_{0\dots M-1}] - \lambda \cdot \operatorname{I}[X_{-M\dots -1}, Z] \right)
\end{equation}
%where $Z$ ranges over all random variables such that the Markov condition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot X_{0\dots M-1} | X_{-M\dots -1}
%\end{equation}
It will be convenient to equivalently rewrite this as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot X_{0\dots M-1} | X_{-M\dots-1}}	\left[\operatorname{H}[X_{0\dots M-1} | Z] + \lambda \cdot\operatorname{I}[ X_{-M\dots-1}, Z]\right]
\end{equation}
where $\operatorname{H}[X_{0\dots M-1} | Z]$ is the \emph{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[X_{0\dots M-1} | Z]$. % is the reduction in prediction loss when predicting

When deploying such a predictive code $Z$, two components have to be computed: a distribution $P(Z|X_{-M\dots -1})$ that \emph{encodes} past observations into a code $Z$, and a distribution $P(X_{0\dots M-1}|Z)$ that \emph{decodes} the code $Z$ into predictions about the future.

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|X_{-M\dots -1})
\end{equation}
defined using an \emph{encoder} $\phi$, which is expressed in some family of function approximators that transduces an observation sequence $X_{-M\dots -1}$ into the parameters of the distribution $P_\phi(\cdot|X_{-M\dots -1})$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(X_{0\dots M-1}|Z) &= \frac{P(X_{0\dots M-1}, Z)}{P(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P_\phi(X_{0\dots M-1}, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(X_{0\dots M-1}|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[X_{-M\dots -1}, Z]$ of the code $Z \sim P_\phi(Z|x_{-M\dots -1})$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

\begin{proposition}
The loss incurred when predicting the future from $Z$ via $\psi$ upper-bounds the true conditional entropy of the future given $Z$, when predicting using the exact decoding distribution (\ref{eq:cond-fut}):
\begin{equation}\label{ineq1}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(X_{0\dots M-1} | Z)\right] \geq \operatorname{H}[X_{0\dots M-1}|Z]
\end{equation}
Furthermore, equality is attained if and only if $P_\psi(X_{0\dots M-1}|Z) = P(X_{0\dots M-1}|Z)$. % (as defined in~(\ref{eq:cond-fut})).
\end{proposition}

>>>>

\newif \ifcomment
\commenttrue

@@

\DeclarePairedDelimiterX{\infdivx}[2]{(}{)}{%
  #1\;\delimsize\|\;#2%
}
\newcommand{\infdiv}{D\infdivx}
\newcommand{\future}{\overrightarrow{X}}
\newcommand{\past}{\overleftarrow{X}}
\newcommand{\key}{\textbf}
\DeclarePairedDelimiter{\norm}{\lVert}{\rVert}
\newcommand{\finitefuture}{X_{0\dots M-1}}
\newcommand{\finitepast}{X_{-M\dots -1}}

@@

\rljf{Small formatting changes I made: \\
* I prefer bold over italics when introducing terms, because when someone is visually scanning the paper looking for a definition, bold stands out much better. \\
* Punctuation should be attached to equations as if they were part of the text (so sometimes periods and commas after equations) \\
* no spaces between m-dashes---like this \\
Also, something crazy is going on with the line numbers!
}

@@

In Section~\ref{sec:background}, we formally introduce predictive rate distortion, and discuss related notions of predictive complexity.
In Section~\ref{sec:oce}, we describe the prior method of Optimal Causel Estimation (OCE).
In Section~\ref{sec:nprd}, we describe our method NPRD.
In Section~\ref{sec:experiments}, we validate NPRD on processes whose predictive rate distortion is analytically known, showing that it finds essentially optimal predictive models.
In Section~\ref{sec:language}, we apply NPRD to data from five languages, providing the first estimates of predictive rate-distortion of natural language.

@@

We consider stationary discrete-time stochastic processes $(X_t)_{t \in \mathbb{Z}}$, taking values in a state space $S$.
Given a reference point in time, say $T=0$, we are interested in the problem of predicting the future of $X_{t\geq 0} = (X_0, X_1, X_2, ...)$ from the past $X_{t< 0} = (..., X_{-2}, X_{-1})$.
In general---unless the observations $X_t$ are independent--- predicting the future of the process accurately will require taking into account the past observations.
There is a tradeoff between the accuracy of prediction, and how much information about the past is being taken into account.
On one extreme, not taking the past into account at all, one will not be able to take advantage of the dependencies between past and future observations.
On the other extreme, considering the entirety of the past observations $X_{t \leq 0}$ can require storing large and potentially unbounded amounts of information.
This tradeoff between information storage and prediction accuracy is referred to as \key{Predictive Rate-Distortion} (PRD) \citep{marzen-predictive-2016}.
The term \key{rate} refers to the amount of past information being taken into account, while \key{distortion} refers to the degradation in prediction compared to optimal prediction from the full past.

The problem of predictive rate-distortion has been formalized by a range of studies.
A principled and general formalization is provided by applying the Information Bottleneck idea \citep{tishby-information-1999, still-optimal-2010,marzen-predictive-2016}:
We will write $\past$ for the past $X_{< 0}$, and $\future$ for the future $X_{\geq 0}$, following \cite{marzen-predictive-2016}.
We consider random variables $Z$, called \key{codes}, that summarize the past and are used by the observer to predict the future.
Formally, $Z$ needs to be independent of the future $\future$ conditional on the past $\past$: in other words, $Z$ does not provide any information about the future except what is contained in the past. Symbolically:
\begin{equation}\label{eq:markov}
	Z \bot \future | \past.
\end{equation}
This is equivalent to the requirement that $Z \rightarrow \past \rightarrow \future$ be a Markov chain.
This formalizes the idea that the code is computed by an observer from the past, without having access to the future.
Predictions are then made based only on $Z$, without additional access to the past $\past$.

The rate of the code $Z$ is the mutual information between $Z$ and the past: $\operatorname{I}[Z, \past]$.
By the Channel Coding Theorem, this describes the channel capacity that the observer requires in order to transform past observations into the code $Z$.

The distortion is the loss in predictive accuracy when predicting from $Z$, relative to optimal prediction from the full past $\past$.
In the Information Bottleneck formalization, this is equivalent to the amount of mutual information between past and future that is \emph{not} captured by $Z$ \citep{harremoes2007information}: %\rljf{Cite the paper "how to choose a good distortion measure"}
\begin{equation}
	\operatorname{I}[\past, \future|Z].
\end{equation}
Due to the Markov condition, the distortion measure satisfies the relation
\begin{equation}
	\operatorname{I}[\past, \future|Z] = \operatorname{I}[\past, \future] - \operatorname{I}[Z, \future],
\end{equation}
i.e., it captures how much less information $Z$ carries about the future $\future$ compared to the full past $\past$.
For a fixed process $(X_t)_t$, choosing $Z$ to minimize the distortion is equivalent to maximizing the mutual information between the code and the future:
\begin{equation}\label{eq:predictiveness}
\operatorname{I}[Z, \future]
\end{equation}
We will refer to~(\ref{eq:predictiveness}) as the \key{predictiveness} of the code $Z$.

The rate-distortion tradeoff then chooses $Z$ to %minimize rate at bounded distortion:
%$$\min_{Z : \operatorname{I}[\past, \future|Z] \leq d} \operatorname{I}[Z, \past]$$
%, or -- equivalently --
 minimize distortion at bounded rate:
\begin{equation}
\min_{Z :  \operatorname{I}[\past, Z] \leq d}  \operatorname{I}[\past, \future|Z] 
\end{equation}
or -- equivalently -- maximize predictiveness at bounded rate:
\begin{equation}
\max_{Z :  \operatorname{I}[\past, Z] \leq d}  \operatorname{I}[Z, \future].
\end{equation}
Equivalently, for each  $\lambda \geq 0$, we study the problem 
\begin{equation}\label{eq:ib}
	\max_{Z} \left( \operatorname{I}[Z, \future] - \lambda \cdot \operatorname{I}[ \past, Z] \right),
\end{equation}
where the scope of the maximization is the class of all random variables $Z$ such that $Z \rightarrow \past \rightarrow \future$ is a Markov chain. %, that is, $Z$ is independent of the future given the past.

The objective (\ref{eq:ib}) is equivalent to the Information Bottleneck \citep{tishby-information-1999}, applied to the past and future of a stochastic process.
The coefficient $\lambda$ indicates how strongly a high rate $\operatorname{I}[\past, Z]$ is penalized; higher values of $\lambda$ result in lower rates and thus lower values of predictiveness.

The largest achievable predictiveness $\operatorname{I}[Z, \future]$ is equal to $\operatorname{I}[\past, \future]$, which is known as the \key{excess entropy} of the process~\citep{feldman-synchronizing-2004}.
Due to the Markov condition~(\ref{eq:markov}) and the Data Processing Inequality, predictiveness of a code $Z$ is always upper-bounded by the rate:
\begin{equation}
\operatorname{I}[Z, \future] \leq \operatorname{I}[\past, Z] 
\end{equation}
As a consequence, when $\lambda \geq 1$, then (\ref{eq:ib}) is always optimized by a trivial $Z$ with zero rate and zero predictiveness.
When $\lambda = 0$, any lossless code optimizes the problem.
Therefore, we will be concerned with the situation where $\lambda \in (0,1)$.

%
%
%Predictive rate-distortion describes the tradeoff between the cost of representing features of the past, and the quality of predictions in the future.
%Let $(X_t)$ be a stationary process.
%We consider feature representations $Z$ of the past $\past$ that are used to predict the future $\future$.
%There is a tradeoff between the complexity of $Z$ and the success of predicting the future:
%More complex representations of the past can reduce future prediction loss -- up to a certain limit given by the entropy rate of the process. %This limit corresponds to perfect modeling, and results in Crutchfield and colleagues' notion of complexity.
%
%The complexity of the representation $Z$, or the \emph{rate}, is formalized as the mutual information between the past and the representation $Z$: $\operatorname{I}[Z, \past]$.
%Different distortion measures could be used as measures of prediction loss.
%CITE maximize the entropy, or, equivalently, minimize the mutual information, between $Z$ and the future: $\operatorname{I}[Z, \past]$.
%
%Thus, for any process $(X_t)$, we obtain a rate-distortion tradeoff curve.
%The achievable region is populated by the random variables $Z$ such that $Z \rightarrow \left(\past\right) \rightarrow \left(\future\right)$ is a Markov chain.
%The points on the tradeoff curve are obtained by solving, for each $\beta \in [0,1]$, the problem
%\begin{equation}\label{eq:ib}
%	\max_{Z} \left( \operatorname{I}[\future, Z] - \beta \cdot \operatorname{I}[Z, (\past)] \right)
%\end{equation}
%where the scope of the maximization is the class of all random variables $Z$ such that $Z - \left(\past\right) - \left(\future\right)$ is a Markov chain, that is, $Z$ is independent of the future given the past.
%This is exactly the Information Bottleneck \cite{tishby-information-1999}, applied to the past and future of a stochastic process.
%
%

@@

Predictive Rate-Distortion is closely related to Statistical Complexity and the $\epsilon$-machine \citep{crutchfield-inferring-1989,Grassberger1986}. %\rljf{cite Grassberger 1986}
Given a stationary process $X_t$, its \key{causal states} are the equivalence classes of semi-infinite pasts $\past$ that induce the same conditional probability over semi-infinite futures $\future$:
Two pasts $\past, \past^\prime$ belong to the same causal state if and only if $P(x_{1...M}|\past) = P(x_{1...M}|\past^\prime)$ holds for all finite sequences $x_{0...M}$ ($M \in \mathbb{N}$).
Note that this definition is not measure-theoretically rigorous; such a treatment is provided by~\citet{lohr-properties-2009}.

The causal states constitute the state set of a a Hidden Markov Model (HMM) for the process, referred to as the \key{$\epsilon$-machine}  \citep{crutchfield-inferring-1989}.
The \emph{Statistical Complexity} of a process is the state entropy of the $\epsilon$-machine.
Statistical complexity can be computed easily if the $\epsilon$-machine is analytically known, but estimating statistical complexity empirically from time series data is very challenging and seems to at least require additional assumptions about the process~\citep{clarke2003application}.

@@

There are related models that represent past observations by extracting those features that are relevant for prediction.
Predictive State Representations \citep{singh-learning-2003,singh-predictive-2004} and Observable Operator Models \citep{jaeger1998discrete} encode past observations as sets of predictions about future observations.
\citet{rubin2012trading} study agents that trade the cost of representing information about the environment against the reward they can obtain by choosing actions based on the representation.
Relatedly, \citet{still-information-2014} introduces a Recursive Past Future Information Bottleneck where past information is compressed repeatedly, not just at one reference point in time.

As discussed in Section~\ref{sec:complexity}, estimating predictive rate-distortion is related to the problem of estimating statistical complexity.
\citet{clarke2003application} and \citet{still-optimal-2010} consider the problem of estimating statistical complexity from finite data.
While statistical complexity cannot be identified from finite data in general, \citet{clarke2003application} introduce certain restrictions on the underlying process that make this more feasible.

@@

The main prior method for estimating predictive rate-distortion from data is Optimal Causal Estimation (OCE, \citet{still-optimal-2010}).
This method approximates predictive rate-distortion using two approximations:
First, it replaces semi-infinite pasts and futures with bounded-length contexts $(X_{-M \dots -1}), (X_{0 \dots M-1})$:
\begin{equation}\label{eq:ib-oce}
	\max_{Z} \left( \operatorname{I}[X_{-M\dots -1}, Z] - \lambda \cdot \operatorname{I}[Z, \finitefuture] \right).
\end{equation}
Second, it estimates information measures directly from the observed counts of $(X_{-M\dots -1}), (\finitefuture)$ using the plug-in estimator of mutual information. %\rljf{Should be \dots instead of commas?}
With such an estimator, the problem in (\ref{eq:ib}) can be solved using a variant of the Blahut-Arimoto algorithm \citep{tishby-information-1999}, obtaining an encoder $P(Z|X_{-M \dots -1})$ that maps each observed past sequence $X_{-M...-1}$ to a distribution over a (finite) set of code words $Z$.
%From this, one obtains a corresponding decoder $P(X_{0...M}|Z)$ via Bayes' Rule.

Two main challenges have been noted in prior work:
First, solving the problem for a finite empirical sample leads to overfitting, overestimating the amount of structure in the process.
\citet{still-optimal-2010} address this by subtracting an asymptotic correction term that becomes valid in the limit of large $M$ and $\lambda \rightarrow 0$, when the codebook $P(Z|\past)$ becomes deterministic, and which allows them to select a deterministic codebook of an appropriate complexity.
This leaves open how to obtain estimates outside of this regime, when the codebook can be far from deterministic.

@@

A further challenge is that OCE is infeasible if the number of required codewords is too large, again because OCE requires constructing a matrix whose rows and columns are indexed by the codewords and observed sequences.
Given that storing and manipulating matrices greater than $10^5 \times 10^5$ is currently not feasible, a setting where $\operatorname{I}[\past, Z] > \log 10^5 \approx 11.5$ cannot be captured with OCE.

@@

We now introduce our method, Neural Predictive Rate Distortion (NPRD), to address the limitations of OCE, by using parametric function approximation:
Whereas OCE constructs a codebook mapping between observed sequences and codes, we use general-purpose function approximation estimation methods to compute the representation $Z$ from the past and to estimate a distribution over future sequences from $Z$.
In particular, we will use recurrent neural networks, which are known to provide good models of sequences from a wide range of domains; our method will be applicable to other types of function approximators also.
%The parameters of the function approximators are optimized on the observed pairs of pasts and futures.
%When deploying the resulting mapping on unseen data, the function approximator can be used to generate appropriate representations $Z$ even for sequences that did not occur in the initial data sample.

@@

As in OCE (Section~\ref{sec:oce}), we assume that a set of finite sample trajectories $x_{-M} \dots x_{M-1}$ is available, and we aim to compress pasts of length $M$ to predict futures of length $M$. To carry this out, we restrict the PRD objective~(\ref{eq:ib}) to such finite-length pasts and futures: \rljf{I think it would be a lot clearer to use the notation from other papers where it's like $\past$ except the arrow points to $M$. It's hard visually looking at the equations to distinguish $-M\dots-1$ from $0\dots M-1$, and that makes it harder to read.}
\begin{equation}\label{eq:ib-oce-repeat}
	\max_{Z : Z \bot \finitefuture | X_{-M\dots-1}} \left( \operatorname{I}[Z, \finitefuture] - \lambda \cdot \operatorname{I}[X_{-M\dots -1}, Z] \right).
\end{equation}
%where $Z$ ranges over all random variables such that the Markovcondition~(\ref{eq:markov}) holds:
%\begin{equation}
%    Z\bot \finitefuture | X_{-M\dots -1}
%\end{equation}
It will be convenient to equivalently rewrite (\ref{eq:ib-oce-repeat}) as
\begin{equation}\label{eq:prd-loss}
	\min_{Z : Z \bot \finitefuture | X_{-M\dots-1}}	\left[\operatorname{H}[\finitefuture | Z] + \lambda \cdot\operatorname{I}[ X_{-M\dots-1}, Z]\right],
\end{equation}
where $\operatorname{H}[\finitefuture | Z]$ is the \key{prediction loss}. Note that minimizing prediction loss is equivalent to maximizing predictiveness $\operatorname{I}[\finitefuture | Z]$. % is the reduction in prediction loss when predicting

When deploying such a predictive code $Z$, two components have to be computed: a distribution $P(Z|X_{-M\dots -1})$ that \key{encodes} past observations into a code $Z$, and a distribution $P(\finitefuture|Z)$ that \key{decodes} the code $Z$ into predictions about the future.

Let us assume that we already have some encoding distribution
\begin{equation}
Z \sim P_\phi(Z|X_{-M\dots -1}),
\end{equation}
where $\phi$ is the encoder, expressed in some family of function approximators. The encoder transduces an observation sequence $X_{-M\dots -1}$ into the parameters of the distribution $P_\phi(\cdot|X_{-M\dots -1})$.
From this encoding distribution, one can obtain the optimal decoding distribution over future observations via Bayes' rule: %, using the Markov condition: 
\begin{equation}
	\label{eq:cond-fut}
	\begin{split}
	P(\finitefuture|Z) &= \frac{P(\finitefuture, Z)}{P(Z)} \\
	&= \frac{\E_{X_{-M \dots -1}} P(\finitefuture, Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
	&=^{(*)} \frac{\E_{X_{-M \dots -1}}   P(\finitefuture|X_{-M\dots-1})     P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})} \\
%	&= \frac{\E_{X_{-M \dots M-1}} P_\phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} P_\phi(Z|X_{-M \dots -1})}
	\end{split}
\end{equation}
where $(*)$ uses the Markov condition.
However, neither of the two expectations in the last line of~(\ref{eq:cond-fut}) is tractable, as they require summation over exponentially many sequences, and algorithms (e.g., dynamic programming) to compute this sum efficiently are not available in general.
For a similar reason, the rate $\operatorname{I}[X_{-M\dots -1}, Z]$ of the code $Z \sim P_\phi(Z|x_{-M\dots -1})$ is also generally intractable.
%If $\phi$ is given by a powerful function approximating family such as recurrent neural networks,

@@

\begin{proposition}
The loss incurred when predicting the future from $Z$ via $\psi$ upper-bounds the true conditional entropy of the future given $Z$, when predicting using the exact decoding distribution (\ref{eq:cond-fut}):
\begin{equation}\label{ineq1}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(\finitefuture | Z)\right] \geq \operatorname{H}[\finitefuture|Z].
\end{equation}
Furthermore, equality is attained if and only if $P_\psi(\finitefuture|Z) = P(\finitefuture|Z)$. % (as defined in~(\ref{eq:cond-fut})).
\end{proposition}

\begin{proof}
By Gibbs' inequality:
\begin{align*}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(\finitefuture | Z)\right] & \geq -	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P(\finitefuture | Z)\right]\\
	&= \operatorname{H}[\finitefuture|Z].
\end{align*}
\end{proof}
